---
title: "CAP-6619 Homework 2, question 5 - lines for each iteration"
author: Christian Garbin
date: Fall 2018
output: html_notebook
---

Shows the lines that represent the different values of the weights that neuralnet() chose during
training.

The goal is to compare the results from each repetition as an attempt to understand better how
`neuralnet()` works.

# Setup the environment

```{r error=TRUE, warning=TRUE}
# Always start with a clean environment to avoid subtle bugs
rm(list = ls())

# To get repeatable results with random numbers (easier to debug multiple runs)
# Note: only works when running the notebook from the top. When running cells individually,
# add this line to each cell.
set.seed(1234) 

setwd("~/fau/cap6619/assignments/assignment2")
```

# Load data

```{r error=TRUE, warning=TRUE}
training_data <- read.table("nine.instances.txt",sep = ",",header = T)
```

# Train the network

```{r error=TRUE, warning=TRUE}
library(neuralnet)
nn <- neuralnet(Label~X1+X2,training_data,hidden = 2,rep = 5)
```

# Plot lines for each iteration

#### Helper functions

Helper function for intercept and slope calculation.

```{r}
findinterceptslope <- function(weights) {
  slope <- weights[2]/weights[3]*(-1)
  intercept <- weights[1]/weights[3]*(-1)
  return(list(slope = slope, intercept = intercept))
}
```

Helper function to plot the lines of the given iteration.

```{r error=TRUE, warning=TRUE}
plot_lines <- function(repetition) {
  # Weights for one repetition (all neurons, including output neuron)
  weights <- nn$weights[repetition]

  # Weights for the first hidden neuron (at the top)
  weights_top = weights[[1]][[1]][1:3,1]

  # Weights for the second hidden neuron (at the bottom)
  weights_bottom = weights[[1]][[1]][1:3,2]
  
  # Slope and intercept for the hidden neurons
  slope_intercept_top = findinterceptslope(weights_top)
  slope_intercept_bottom = findinterceptslope(weights_bottom)

  # Uncomment to see the raw values
  #print(repetition)
  #print(weights_top)
  #print(weights_bottom)
  #cat(slope_intercept_top$slope, slope_intercept_top$intercept, "\n")
  #cat(slope_intercept_bottom$slope, slope_intercept_bottom$intercept, "\n\n")

  # Plot training data and lines (hyperplanes) from the hidden neurons
  abline(slope_intercept_top$intercept,slope_intercept_top$slope,col = "green",lty = 2)
  abline(slope_intercept_bottom$intercept,slope_intercept_bottom$slope,col = "blue",lty = 2)
}
```

Plots data and lines for the given repetition.

```{r error=TRUE, warning=TRUE}
plot_repetition <- function(repetition) {
  header <- paste(repetition, "e=", round(nn$result.matrix["error",repetition],5))

  # xlim and ylim were chose by trial-and-error to allow better visual comparison of the lines  
  plot(training_data$X1,training_data$X2,col = c("black","red")[training_data$Label + 1],
       xlim = c(-1.3,1.3),ylim = c(-1.3,1.3), main = header)

  plot_lines(repetition)
}
```

#### Side-by-side for comparison

See next session for individual plots.

```{r error=TRUE, warning=TRUE}
par(mfrow = c(1,5))
plot_repetition(1)
plot_repetition(2)
plot_repetition(3)
plot_repetition(4)
plot_repetition(5)
```

#### Individual plots

```{r error=TRUE, warning=TRUE}
plot_repetition(1)
```

```{r error=TRUE, warning=TRUE}
plot_repetition(2)
```

```{r error=TRUE, warning=TRUE}
plot_repetition(3)
```

```{r error=TRUE, warning=TRUE}
plot_repetition(4)
```

```{r error=TRUE, warning=TRUE}
plot_repetition(5)
```
